Acceleration of Bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-11 trunk/LateX/paper8.tex
more minor changes

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel. A formal expression is given below:

@@

On the GPU, our pair-symmetric algorithm outperformed our naive algorithm for every set of tests. For example, Figure~\ref{fig:hw10} shows the pair-symmetric algorithm's best timing to be roughly 75\% of the naive algorithm's best timing for processing a 3000 x 3000 image using a 21 x 21 pixel filter window. The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size. In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm. Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased. Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.

While our tests returned a maximum 25\% reduction in execution time, our pair-symmetric algorithm has the potential to reach a near 50\% reduction in execution time. Furthermore, the benchmarks of our pair-symmetric algorithm reveal minor irregularities in the trajectory of performance derived from execution timings versus respective incremental adjustments of parameter values such as CUDA block and grid dimensions. The main cause of both of these observations is shared memory. To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time. Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings. To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

>>>>

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel. A formal expression is given below:

@@

On the GPU, our pair-symmetric algorithm outperformed our naive algorithm for every set of tests. For example, Figure~\ref{fig:hw10} shows the pair-symmetric algorithm's best timing to be roughly 75\% of the naive algorithm's best timing for processing a 3000 x 3000 image using a 21 x 21 pixel filter window. The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size. In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm. Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased. Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect our tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.

While our tests returned a maximum 25\% reduction in execution time, our pair-symmetric algorithm has the potential to reach a near 50\% reduction in execution time. Furthermore, the benchmarks of our pair-symmetric algorithm reveal minor irregularities in the trajectory of performance derived from execution timings versus respective incremental adjustments of parameter values such as CUDA block and grid dimensions. The main cause of both of these observations is shared memory. To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time. Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., reducing timings by no more than than 25\%. To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code, CUDA's quasi-assembly language, of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.
